Composable data visualizations

### 2020-06-12 wip-compost/paper.tex
Fix typos

Now we can define the besic elements of the Compost domain-specific langauge.
A chart element is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
A primitive shape can be a text label, a line connecting a list of points, a filled polygon
defined by a list of points or a bubble at a given point with a given width and height.
The position of points is specified by \X and \Y coordinates, which can be either categorical or
continuous values. For text, line, polygon and bubble, we also include a parameter $\gamma$
that specifies the element color. The width and height of a bubble is given in pixels rather than
in domain units.

@@

We introduce the Compost domain-specific language as untyped, but there are some obvious ways in
which types can make composing charts in Compost safer. First, a type representing a shape could
specify whether the \X and \Y axes represent categorical or continous values. This would rule out
mixing of different values on a single scale and guarantee that the \ident{union} operation,
sketched in the previous section, is never called in a way leading to an undefined result.
Second, the type of values mapped to an axis could be further annotated with units of measure
\cite{units}. Using the F\# notation where $n\langl u\rangl$ is a number $n$ with unit $u$,
an axis containing a value $\kvd{cont}~\num{317}\langl \ident{mp}\rangl$ would then be incompatible
with an axis containing a value $\kvd{cont}~\num{1.32}\langl \ident{gbp}/\ident{usd}\rangl$.

@@

The automatic scale inference often results in scales where the maximum is a non-round number.
This leads to charts that fully utilize the available space, but may not be easy to read.
The first two primitives, shown in Figure~\ref{fig:control} (left) allow the chart designer
to adjust the automatically inferred range of scales.
The oprations can be applied to either the \X scale or the \Y scale, which is indicated by the
$x/y$ subscript. The \kvd{roundScale} primitive takes the inferred \X or \Y scale of the shape $s$
and, if it is a continuous scale, rounds its minimal and maximal values to a ``nice'' number.
For example, if a continuous scale has minimum $\num{0}$ and maximum $\num{365}$, the resulting
scale would have a maximum $\num{400}$. For categorical scale, the operation does not have any effect.
The \kvd{explicitScale} operation replaces the inferred scale with an explicitly
provided scale (the type of the inferred scale has to match with the type of the explicitly given
scale). For example, the chart in Figure~\ref{fig:simple} (right) is constructed using the
following code (reusing the \ident{labour} and \ident{conservative} variables defined earlier):
%
\begin{equation*}
\begin{array}{l}
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{roundScale}_y~(\kvd{explicitScale}_x~(\kvd{categorical}~[\,\strf{Labour},\strf{Conservative}\,]),\\
\qquad \kvd{overlay}~[~\ident{labour}, \ident{conservative}~]~)))\\
\end{array}
\end{equation*}
\vspace{-0.75em}

@@

The most interesting primitive for controlling scale composition
is $\kvd{nest}_{x/y}$. The behaviour is similar to that of the SVG viewport \cite{svg}.
It takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments and
nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$.
When inferring scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
constructed from the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
Here, a chart with a continous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.

@@

The nesting of scales can be used in a variety of ways. For example, to nest a scatter plot
showing individual data points inside a bar of a histogram, we would use $\kvd{cat}~\strf{ABC}, 0$
and $\kvd{cat}~\strf{ABC}, 1$ as the points that define the start and the end of the region. A simpler
use case for the combinator is showing multiple charts in a single view. For example, the motivating example
in Figure~\ref{fig:charts} (right) comapres aligned line charts of exchange rates for two
different currencies. Assuming \ident{gbpusd} and \ident{gbpeur} are lists containing days as \X
values and exchange rates as \Y values, we can construct a simple chart with two line charts,
shown in Figure~\ref{fig:lines} (left), using:
%
\begin{equation*}
\begin{array}{l}
\kvd{overlay}~[~
\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpusd}))))\\
\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpeur}))))~]\\
\end{array}
\end{equation*}

@@

\section{Standard charts: Definining new abstractions}
\label{sec:abstractions}

@@

A more complex chart that can be composed using the Compost primitives is pairplot from the seaborn
library \cite{seaborn}. Pairplot visualizes pairwise relationships between features of a dataset.
An example using three features (sepal width, petal width, petal length) from the Iris dataset is shown
in Figure~\ref{fig:standard} (right).
A pairplot draws a grid of charts, each visualizing the relationship between two numerical features. For distinct
features, pairplot shows a scatter plot using one fature for \X values and the other for \Y values.
When the fatures are the same (the diagonal), it draws a histogram of the feature values.
A categorical feature can be used to determine the color of dots in the scatter plots.

@@

\vspace{-0.5em}
\noindent
As before, \kvd{nest} is essential for composing individual charts. Here, the points that determine
the locations of individual charts are categorical values defined by the attributes of the dataset.
The choice between two possible nested charts is made using the host language \fkvd{if} construct.
Scatter plots are generated by overlaying bubbles with \X and \Y coordinates obtained using $\ident{get}~x~r$
and $\ident{get}~y~r$. Histograms are composed from filled shapes. To obtain their locations, we use
a helper function $\ident{bins}~x~\ident{data}$, which returns a list of bins specified by a tripple
consisting of a lower and an upper range $x_1, x_2$ and the count $y$.

@@

\vspace{-0.5em}
\noindent
The first case handles the \ident{Moving} event, which replaces the first component of the
state tuple, i.e.~a flag indicating whether a mouse button is down.
The next two cases handle the \ident{Update} event. The event carries two values, $p, v$, which
represent the party (which bar the user is dragging) and the new value (new number of seats).
If the user is currently dragging, we replace the value associated with the party $p$ in \noindent the
list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ingored.

@@

When the user interacts with the visualization created using Compost, the library
translates the coordinates associated with events from pixels to domain-specific
values. In case of the above bar chart, when the user moves a mouse, the function registed
using \kvd{mouseMove} is given a categorical value $\kvd{cat}~p, r$ as the \X coordinate and
a continuous value $\kvd{cont}~v$ as the \Y coordinate. It then takes $p$, which is the name
of the party corresponding to the bar and the value $v$ corresponding to the number of seats
and triggers the $\ident{Update}(p, v)$ event to update the state. The handlers for
\kvd{mouseUp} and \kvd{mouseDown} do not use the coordinates. They simply switch the flag
indicating whether the user is currently dragging or not.

@@

Compost is an open-source library, implmented in the functional language F\#. The full source
code can be found at \urrl{http://github.com/compostjs}.
As is often the case with functional domain-specific languages, the implementation is not difficult
once we find the right collection of basic primitives and the right structure for the implementation.
This largely applies to the Compost library and so we will not go into the implementation details.
It is, however, worth giving an outline of the implmentation structure.

@@

In order to render a shape, we need to know the range of values that should appear on the \X and
\Y axes. This is done by inferring a scale for each of the axes from the individual \X and \Y
coordinates that specify shape locations. As discussed earlier, a scale can be either categorical
(displaying only categorical values) or contiuous (displaying only continuous values). When
inerring scales, we use two helper operations; \ident{union}, discussed earlier, combines two
scales and \ident{singleton} creates a scale from a single coordinate.

>>>>

Now we can define the basic elements of the Compost domain-specific language.
A chart element is represented by the shape $s$ defined in Figure~\ref{fig:shape}.
A primitive shape can be a text label, a line connecting a list of points, a filled polygon
defined by a list of points or a bubble at a given point with a given width and height.
The position of points is specified by \X and \Y coordinates, which can be either categorical or
continuous values. For text, line, polygon and bubble, we also include a parameter $\gamma$
that specifies the element color. The width and height of a bubble is given in pixels rather than
in domain units.

@@

We introduce the Compost domain-specific language as untyped, but there are some obvious ways in
which types can make composing charts in Compost safer. First, a type representing a shape could
specify whether the \X and \Y axes represent categorical or continuous values. This would rule out
mixing of different values on a single scale and guarantee that the \ident{union} operation,
sketched in the previous section, is never called in a way leading to an undefined result.
Second, the type of values mapped to an axis could be further annotated with units of measure
\cite{units}. Using the F\# notation where $n\langl u\rangl$ is a number $n$ with unit $u$,
an axis containing a value $\kvd{cont}~\num{317}\langl \ident{mp}\rangl$ would then be incompatible
with an axis containing a value $\kvd{cont}~\num{1.32}\langl \ident{gbp}/\ident{usd}\rangl$.

@@

The automatic scale inference often results in scales where the maximum is a non-round number.
This leads to charts that fully utilise the available space, but may not be easy to read.
The first two primitives, shown in Figure~\ref{fig:control} (left) allow the chart designer
to adjust the automatically inferred range of scales.
The operations can be applied to either the \X scale or the \Y scale, which is indicated by the
$x/y$ subscript. The \kvd{roundScale} primitive takes the inferred \X or \Y scale of the shape $s$
and, if it is a continuous scale, rounds its minimal and maximal values to a ``nice'' number.
For example, if a continuous scale has minimum $\num{0}$ and maximum $\num{365}$, the resulting
scale would have a maximum $\num{400}$. For categorical scale, the operation does not have any effect.
The \kvd{explicitScale} operation replaces the inferred scale with an explicitly
provided scale (the type of the inferred scale has to match with the type of the explicitly given
scale). For example, the chart in Figure~\ref{fig:simple} (right) is constructed using the
following code (reusing the \ident{labour} and \ident{conservative} variables defined earlier):
%
\begin{equation*}
\begin{array}{l}
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{roundScale}_y~(\kvd{explicitScale}_x~(\kvd{categorical}~[\,\strf{Labour},\strf{Conservative}\,]),\\
\qquad \kvd{overlay}~[~\ident{labour}, \ident{conservative}~]~)))\\
\end{array}
\end{equation*}
\vspace{-0.75em}

@@

The most interesting primitive for controlling scale composition
is $\kvd{nest}_{x/y}$. The behaviour is similar to that of the SVG viewport \cite{svg}.
It takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments and
nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$.
When inferring scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
constructed from the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
Here, a chart with a continuous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.

@@

The nesting of scales can be used in a variety of ways. For example, to nest a scatter plot
showing individual data points inside a bar of a histogram, we would use $\kvd{cat}~\strf{ABC}, 0$
and $\kvd{cat}~\strf{ABC}, 1$ as the points that define the start and the end of the region. A simpler
use case for the combinator is showing multiple charts in a single view. For example, the motivating example
in Figure~\ref{fig:charts} (right) compares aligned line charts of exchange rates for two
different currencies. Assuming \ident{gbpusd} and \ident{gbpeur} are lists containing days as \X
values and exchange rates as \Y values, we can construct a simple chart with two line charts,
shown in Figure~\ref{fig:lines} (left), using:
%
\begin{equation*}
\begin{array}{l}
\kvd{overlay}~[~
\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpusd}))))\\
\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpeur}))))~]\\
\end{array}
\end{equation*}

@@

\section{Standard charts: Defining new abstractions}
\label{sec:abstractions}

@@

A more complex chart that can be composed using the Compost primitives is pairplot from the seaborn
library \cite{seaborn}. Pairplot visualizes pairwise relationships between features of a dataset.
An example using three features (sepal width, petal width, petal length) from the Iris dataset is shown
in Figure~\ref{fig:standard} (right).
A pairplot draws a grid of charts, each visualizing the relationship between two numerical features. For distinct
features, pairplot shows a scatter plot using one feature for \X values and the other for \Y values.
When the features are the same (the diagonal), it draws a histogram of the feature values.
A categorical feature can be used to determine the color of dots in the scatter plots.

@@

\vspace{-0.5em}
\noindent
As before, \kvd{nest} is essential for composing individual charts. Here, the points that determine
the locations of individual charts are categorical values defined by the attributes of the dataset.
The choice between two possible nested charts is made using the host language \fkvd{if} construct.
Scatter plots are generated by overlaying bubbles with \X and \Y coordinates obtained using $\ident{get}~x~r$
and $\ident{get}~y~r$. Histograms are composed from filled shapes. To obtain their locations, we use
a helper function $\ident{bins}~x~\ident{data}$, which returns a list of bins specified by a triple
consisting of a lower and an upper range $x_1, x_2$ and the count $y$.

@@

\vspace{-0.5em}
\noindent
The first case handles the \ident{Moving} event, which replaces the first component of the
state tuple, i.e.~a flag indicating whether a mouse button is down.
The next two cases handle the \ident{Update} event. The event carries two values, $p, v$, which
represent the party (which bar the user is dragging) and the new value (new number of seats).
If the user is currently dragging, we replace the value associated with the party $p$ in \noindent the
list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ignored.

@@

When the user interacts with the visualization created using Compost, the library
translates the coordinates associated with events from pixels to domain-specific
values. In case of the above bar chart, when the user moves a mouse, the function registered
using \kvd{mouseMove} is given a categorical value $\kvd{cat}~p, r$ as the \X coordinate and
a continuous value $\kvd{cont}~v$ as the \Y coordinate. It then takes $p$, which is the name
of the party corresponding to the bar and the value $v$ corresponding to the number of seats
and triggers the $\ident{Update}(p, v)$ event to update the state. The handlers for
\kvd{mouseUp} and \kvd{mouseDown} do not use the coordinates. They simply switch the flag
indicating whether the user is currently dragging or not.

@@

Compost is an open-source library, implemented in the functional language F\#. The full source
code can be found at \urrl{http://github.com/compostjs}.
As is often the case with functional domain-specific languages, the implementation is not difficult
once we find the right collection of basic primitives and the right structure for the implementation.
This largely applies to the Compost library and so we will not go into the implementation details.
It is, however, worth giving an outline of the implementation structure.

@@

In order to render a shape, we need to know the range of values that should appear on the \X and
\Y axes. This is done by inferring a scale for each of the axes from the individual \X and \Y
coordinates that specify shape locations. As discussed earlier, a scale can be either categorical
(displaying only categorical values) or continuous (displaying only continuous values). When
inferring scales, we use two helper operations; \ident{union}, discussed earlier, combines two
scales and \ident{singleton} creates a scale from a single coordinate.
